---
title: "Case study: digit recognition with scBubbletree"
author: "SK"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = F,
                      comment = F, 
                      warning = F, 
                      message = F)
```

# Background
While `r Biocpkg("scBubbletree")` is primarily designed for the analysis of 
scRNA-seq data, in this vignette we show that this package can be used as a 
general purpose tool for visual exploration of "big data".

To run this vignette we need to load a few R-packages:

```{r}
library(scBubbletree)
library(ggplot2)
library(ggtree)
library(patchwork)
```


# Data

We will explore the **digits** dataset. The data contains 64 feature (8x8 
pixel values in the range 0 to 16) for 5620 images.

```{r, echo=T}
# .. _digits_dataset:
# 
# Optical recognition of handwritten digits dataset
# --------------------------------------------------
# 
# Data Set Characteristics:
# 
#     :Number of Instances: 5620
#     :Number of Attributes: 64
#     :Attribute Information: 8x8 image of integer pixels in the range 0..16.
#     :Missing Attribute Values: None
#     :Creator: E. Alpaydin (alpaydin '@' boun.edu.tr)
#     :Date: July; 1998
# 
# This is a copy of the test set of the UCI ML hand-written digits datasets
# https://archive.ics.uci.edu/ml/datasets/Optical+Recognition+of+Handwritten+Digits
# 
# The data set contains images of hand-written digits: 10 classes where
# each class refers to a digit.
# 
# Preprocessing programs made available by NIST were used to extract
# normalized bitmaps of handwritten digits from a preprinted form. From a
# total of 43 people, 30 contributed to the training set and different 13
# to the test set. 32x32 bitmaps are divided into nonoverlapping blocks of
# 4x4 and the number of on pixels are counted in each block. This generates
# an input matrix of 8x8 where each element is an integer in the range
# 0..16. This reduces dimensionality and gives invariance to small
# distortions.
# 
# For info on NIST preprocessing routines, see M. D. Garris, J. L. Blue, G.
# T. Candela, D. L. Dimmick, J. Geist, P. J. Grother, S. A. Janet, and C.
# L. Wilson, NIST Form-Based Handprint Recognition System, NISTIR 5469,
# 1994.
# 
# .. topic:: References
# 
#   - C. Kaynak (1995) Methods of Combining Multiple Classifiers and Their
#     Applications to Handwritten Digit Recognition, MSc Thesis, Institute of
#     Graduate Studies in Science and Engineering, Bogazici University.
#   - E. Alpaydin, C. Kaynak (1998) Cascading Classifiers, Kybernetika.
#   - Ken Tang and Ponnuthurai N. Suganthan and Xi Yao and A. Kai Qin.
#     Linear dimensionalityreduction using relevance weighted LDA. School of
#     Electrical and Electronic Engineering Nanyang Technological University.
#     2005.
#   - Claudio Gentile. A New Approximate Maximal Margin Classification
#     Algorithm. NIPS. 2000.
```


```{r}

# dreate temp folder
dir.create(path = "opt_tmp")

# download input data
utils::download.file(
  url = "https://archive.ics.uci.edu/ml/machine-learning-databases/optdigits/optdigits.tra", 
  destfile = "opt_tmp/optdigits.tra")
utils::download.file(
  url = "https://archive.ics.uci.edu/ml/machine-learning-databases/optdigits/optdigits.tes", 
  destfile = "opt_tmp/optdigits.tes")

# read in digitis
digits <- rbind(as.matrix(read.csv(file = "opt_tmp/optdigits.tra", header = F)),
                as.matrix(read.csv(file = "opt_tmp/optdigits.tes", header = F)))
labels <- digits[,65]
digits <- digits[, -65]

# remove tmp once digitis dataset has been loaded
unlink("opt_tmp/", recursive=TRUE)
```


# scBubbletree workflow
As first input `r Biocpkg("scBubbletree")` uses matrix $A^{n\times f}$ (here 
`A=digits`). We will skip the data processing that are ususal for scRNA-seq 
data, such as normalization, linear dimensionality reduction, etc.

```{r}
# A has n=cells as rows, f=features as columns (e.g. from PCA)
dim(A)
```

The `r Biocpkg("scBubbletree")` workflow performs the following steps:
  
  1. determine the clustering resolution (clusters $k$ or resolution $r$)
  2. graph-based community detection (e.g. Louvain) or k-means clustering
  3. hierarchical organization of clusters (bubbles)
  4. visualization

## 1. determine the clustering resolution
If we use graph-based community detection (**recommended for scRNA-seq**) with
e.g. the Louvain or Leiden method, then we need to find appropriate value for 
the resolution parameter $r$. Otherwise, we can use the simpler k-means 
clustering algorithm in which case we need to find an appropriate value of the 
number of clusters $k$.

### 1.1 determining the number of clusters $k$
How many clusters are there are in the data? Can we guess a reasonable value 
of $k$? We anticipate $k\approx10$ (0, 1, ..., 9)

To perform data-driven search for $k$ using `get_k` we will provide 
the matrix $A$ as input and a vector of $k$s. The output will be the Gap 
statistic and WCSS estimates for each $k$.

Lets run `get_k` now:

```{r, echo = T, eval = T}
b_k <- get_k(B_gap = 5,
           ks = seq(from = 1, to = 20, by = 2),
           x = digits,
           n_start = 100, 
           iter_max = 200,
           kmeans_algorithm = "MacQueen", 
           cores = 4)
```

The Gap statistic and WCSS curves do not have a clear knee (elbow), however,
we see a decrease in the slope of the gap curve for $k>9$.

```{r, echo=T, fig.width=6, fig.height=2.5}
ggplot(data = b_k$gap_stats_summary)+
  geom_point(aes(x = k, y = gap_mean), size = 1)+
  geom_errorbar(aes(x = k, y = gap_mean, ymin = L95, ymax = H95), width = 0.1)+
  ylab(label = "Gap")|
ggplot(data = b_k$wcss_stats_summary)+
  geom_point(aes(x = k, y = wcss_mean), size = 1)+
  ylab(label = "WCSS")+
  scale_y_log10()+
  annotation_logticks(base = 10, sides = "l")
```

### 1.2 determining the resolution parameter $r$
For graph-based community detection we will use the original Louvain method 
and clustering resolution $r$. Higher resolutions lead to more communities and 
lower resolutions lead to fewer communities. We can use the same strategy as 
before to find a reasonable reasonable value of $r$.

Lets use the function `get_r` for data-driven estimation of $r$ based on 
the gap statistic and WCSS. As input we need to provide the matrix $A$ and 
a vector of $r$s. The output will be a gap statistic and WCSS curve for each 
$r$ (or the number of communities $k'$ detected at resolution $r$).

```{r, echo = T}
b_r <- get_r(B_gap = 5,
             rs = 10^seq(from = -4, to = 1.5, by = 0.35),
             x = digits,
             n_start = 20, 
             iter_max = 100,
             algorithm = "original",
             knn_k = 50,
             cores = 4)
```



The gap statistic and WCSS curves have noticeable knees (elbows) at $k'=13$ 
($r=0.71$). Means (points) and 95% confidence intervals are shown for the 
Gap statistic at each $k$ computed using `B_gap`=5 MCMC simulations.

```{r, fig.width=6, fig.height=2.5}
ggplot(data = b_r$gap_stats_summary)+
  geom_point(aes(x = k, y = gap_mean), size = 1)+
  geom_errorbar(aes(x = k, y = gap_mean, ymin = L95, ymax = H95), width = 0.1)+
  ylab(label = "Gap")|
ggplot(data = b_r$wcss_stats_summary)+
  geom_point(aes(x = k, y = wcss_mean), size = 1)+
  ylab(label = "WCSS")+
  scale_y_log10()+
  annotation_logticks(base = 10, sides = "l")
```


A range of resolutions $r\in[0.71, 1.58]$ among the tested $r$s yield $k=13$ 
communities. Lets use $r=0.71$ for clustering.

```{r, fig.width=4, fig.height=3, fig.align='center'}
ggplot(data = b_r$gap_stats_summary)+
    geom_point(aes(x = r, y = k), size = 1)+
    xlab(label = "r")+
    ylab(label = "k'")+
  scale_x_log10()+
  annotation_logticks(base = 10, sides = "b")
```

## 2. clustering and 3. hierarchical grouping
### 2.1 clustering with Louvain
Lets perform clustering with the original Louvain algorithm (function 
FindClusters, R-package `r CRANpkg("Seurat")`) and resolution parameter 
$r=0.71$. We will do clustering with $n_\textit{start} = 20$ random 
starts and a maximum number of iterations $iter_\textit{max} = 100$. 

Steps 2.2 and 3. (hierarchical clustering) are performed next

```{r}
k13_louvain <- get_bubbletree_graph(
  x = digits,
  r = 0.71,
  n_start = 20,
  iter_max = 100,
  algorithm = "original",
  knn_k = 50,
  cores = 1,
  B = 200,
  N_eff = 200,
  round_digits = 1,
  show_simple_count = FALSE)
```


... and plot the bubbletree. We see nearly identical dendrogram as the 
one generated by kmeans clustering. The bubble IDs are different but we 
see similar bubble sizes, topology and branch robustness values.

```{r, fig.width=4, fig.height=5, fig.align='center'}
k13_louvain$tree
```

## 4. visualization

```{r, fig.width=7, fig.height=4, fig.align='center'}
w1 <- get_cat_tiles(btd = k13_louvain,
                    f = as.character(labels),
                    integrate_vertical = FALSE,
                    round_digits = 1,
                    x_axis_name = 'Digits [0-9]',
                    rotate_x_axis_labels = FALSE,
                    tile_text_size = 2.75)

(k13_louvain$tree|w1$plot)+
  patchwork::plot_layout(widths = c(1, 1))
```



```{r}
x <- digits

qs <- c()
for(i in 1:nrow(digits)) {
  cat(i, "\n")
  m <- matrix(data = as.numeric(digits[i, 1:64]), 
              byrow = T, ncol = 8)
  q <- reshape2::melt(m)
  colnames(q) <- c("y", "x", "v")
  q$y <- abs(q$y-9)
  q$i <- i
  q$class <- labels[i]
  q$bubble <- k13_louvain$cluster[i]
  qs <- rbind(qs, q)
}
rm(m, q, i, x)
```


```{r}
q <- aggregate(v~x+y+bubble, data = qs, FUN = mean)
```


# Average digit image in each bubble

Lets compute the average pixel values using the digit data in each bubble and 
'attach' this data to the bubbletree

```{r}
q$bubble <- factor(q$bubble, levels = k13_louvain$tree_meta$label)
g_digit <- ggplot(data = q)+
  facet_wrap(facets = ~bubble, ncol = 2)+
  geom_tile(aes(x = x, y = y, fill = v))+
  theme_bw(base_size = 10)+
  theme(legend.position = "none",
        strip.text.x = element_text(margin = margin(0.015,0,0.015,0, "cm")))+
  scale_fill_gradient(low = "white", high = "black")+
  scale_x_continuous(name = "x pixel", expand = c(0, 0))+
  scale_y_continuous(name = "y pixel", expand = c(0, 0))
```


```{r, fig.width=8, fig.height=6}
(k13_louvain$tree|w1$plot|g_digit)+
  patchwork::plot_annotation(tag_levels = 'A')+
  patchwork::plot_layout(widths = c(1,1.5,1))
```



## Session Info

```{r}
sessionInfo()
```
